---
title: "Intersecting species range with island features"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

This is a demonstration of how to intersect species' ranges with island features.

## Setting things up

To set things up, let's read in some snail localities as well as some island data.

```{R}

library(raster)

#import snail data
yann_data <- read.csv("DistributionYannickus.csv")
#get this ready to use
coordinates(yann_data) <- c("Longitude", "Latitude")
proj4string(yann_data) <- CRS("+init=epsg:4326")

#import island data
galap_data<-shapefile("veg_zones.shp")
# the prj file is missing - but that's ok for now

```
Now let's make a species distribution for the snails. For now, we'll use a minimum convex polygon. I think the rest of the script will work with any range that is a polygon.

```{r}
library(adehabitatHR)

yann_distribution <- mcp(yann_data, percent=100, unout="m2")

# let's see how this looks!
plot(galap_data)
plot(yann_distribution, col="pink", add=TRUE)
# cool.

```

Now, let's try to do an intersection. We will use the "humid" zone of the island of Santa Cruz as an example.

```{r}
library(sp)
library(rgeos)

humid_sc<-galap_data[galap_data$NOMBRE_=="Santa Cruz" & galap_data$ZONE=="humid",]
humid_yann<-intersect(humid_sc, yann_distribution)
plot(yann_distribution)
plot(humid_sc, col = "red", add = TRUE)
plot(humid_yann, col="purple", add=TRUE)

# Note that this works for other zones, too 
# even if things get a little weird
trans_sc<-galap_data[galap_data$NOMBRE_=="Santa Cruz" & galap_data$ZONE=="transition",]
trans_yann<-intersect(trans_sc, yann_distribution)
plot(yann_distribution)
plot(trans_sc, col = "green", add = TRUE)
plot(trans_yann, col="yellow", add=TRUE)

# you can also find things out about the intersection
humid_yann$AREA
humid_yann$PERIMETER
```


